Phylogenetics and the human microbiome.

### 2013-06-05 sbreview.tex
some progress organizing

@@

% % For Sys Bio, uncomment this block and comment the previous block.
% \documentclass[12pt,letterpaper]{article}
% \usepackage{fixltx2e}
% \usepackage{textcomp}
% \usepackage{fullpage}
% \usepackage{amsfonts}
% \usepackage{verbatim}
% \usepackage[english]{babel}
% \usepackage{pifont}
% \usepackage{color}
% \usepackage{setspace}
% \usepackage{lscape}
% \usepackage{indentfirst}
% \usepackage[normalem]{ulem}
% \usepackage{booktabs}
% %\usepackage{nag}
% \usepackage{natbib}
% %\usepackage{bibtex}
% \usepackage{float}
% \usepackage{latexsym}
% %\usepackage{hyperref}
% \usepackage{url}
% %\usepackage{html}
% \usepackage{hyperref}
% \usepackage{epsfig}
% \usepackage{graphicx}
% \usepackage{amssymb}
% \usepackage{amsmath}
% \usepackage{bm}
% \usepackage{array}
% %\usepackage{mhchem}
% \usepackage{ifthen}
% \usepackage{caption}
% \usepackage{hyperref}
% %\usepackage{xcolor}
% \usepackage{amsthm}
% \usepackage{amstext}
% \linespread{1.66}
% \raggedright
% \setlength{\parindent}{0.5in}
% \setcounter{secnumdepth}{0}
% \pagestyle{empty}
% \renewcommand{\section}[1]{%
% \bigskip
% \begin{center}
% \begin{Large}
% \normalfont\scshape #1
% \medskip
% \end{Large}
% \end{center}}
% \renewcommand{\subsection}[1]{%
% \bigskip
% \begin{center}
% \begin{large}
% \normalfont\itshape #1
% \end{large}
% \end{center}}
% \renewcommand{\subsubsection}[1]{%
% \vspace{2ex}
% \noindent
% \textit{#1.}---}
% \renewcommand{\tableofcontents}{}
% \bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
% \newcommand{\forarxiv}[1]{}
% \newcommand{\notforarxiv}[1]{#1}

@@

% theorems, etc
\newtheorem{lem}{Lemma}
\newtheorem{cor}{Corollary}
\newtheorem{prop}{Proposition}
\newtheorem{thm}{Theorem}
\newtheorem{prob}{Problem}
\newtheorem{defn}{Definition}
\newtheorem{obs}{Observation}
\newtheorem{alg}{Algorithm}

@@

The human microbiome is the collection of cells that live inside and on humans.
The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.

@@

Starter paragraph about the scope of human microbiome research.
Our idea is to highlight areas that may be of particular interest for the phylogenetics community by potentially having wider applications or by being areas for future exploration.

Terminology: survey versus metagenome.
Definition of a single sample.

The human microbiome is.
Virome.
Has been opened up by high-throughput sequencing.
Non-culturable organisms.

This sequencing has shown that the microbiome contains a diverse collection of organisms, has remarkable metabolic potential.
Number of species.
\cite{qin2010human}
% To understand the impact of gut microbes on human health and well-being it is crucial to assess their genetic potential. Here we describe the Illumina-based metagenomic sequencing, assembly and characterization of 3.3 million non-redundant microbial genes, derived from 576.7 gigabases of sequence, from faecal samples of 124 European individuals. The gene set, ~150 times larger than the human gene complement, contains an overwhelming majority of the prevalent (more frequent) microbial genes of the cohort and probably includes a large proportion of the prevalent human intestinal microbial genes. The genes are largely shared among individuals of the cohort. Over 99% of the genes are bacterial, indicating that the entire cohort harbours between 1,000 and 1,150 prevalent bacterial species and each individual at least 160 such species, which are also largely shared. We define and describe the minimal gut metagenome and the minimal gut bacterial genome in terms of functions present in all individuals and most bacteria, respectively.

@@

Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.

Inter-individual versus intra-individual variation \cite{hmp2012structure}

@@

@@

@@

obesity
turnbaugh papers
http://www.nature.com/nature/journal/v444/n7122/abs/nature05414.html
http://europepmc.org/abstract/MED/18407065/reload=0;jsessionid=Ea4YLfXjd7Cfpx5Hvafq.2
amish study
http://precedings.nature.com/documents/4957/version/1

Data products
reference genomes
HMP DACC.

Core microbiome
\cite{turnbaugh2008core}

@@

\cite{costello2009bacterial}
Bacterial community variation in human body habitats across space and time

\cite{ley2006microbial}
Microbial ecology: human gut microbes associated with obesity

@@

Phylogenetics is primarily used in this realm as a framework with which data can be organized.
It is common for researchers to equate 16s phylogeny with taxonomy.
Trait evolution methods are not common.

@@

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
Bergey's taxonomy is supposed to be.
GreenGenes.
tax2tree.
SILVA.
%The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.

@@

Curation of HM databases using phylogenetics?

@@

\cite{kreig1984bergey}
Bergey's Manual of Systematic Bacteriology.

\cite{dalevi2007automated}
Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool

\cite{desantis2006greengenes}
{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}

@@

Eisen darling paper on extractability.

@@

\subsection{Inference of horizontal gene transfer}

@@

@@

\cite{caporaso2010qiime}

@@

@@

@@

@@

What we can expect next.

Field is still not yet mature.

>>>>

% morgan et al extraction
% PyNAST

@@

@@

@@

The human microbiome is the collection of microbes that live inside of and on the outside of humans.
The Human Microbiome Project (HMP) and the Metagenomics of the Human Intestinal Tract (MetaHIT) consortia have advanced human microboime research by generating thousands of 16s surveys and terabases of metagenomic data on the human microbiome, as well as funding bioinformatics and statistical development.
In this paper I review the impact of phylogenetics and tree-thinking on the methods used in the analysis of this data, as well as describing current challenges for phylogenetics coming from this type of work.

@@

The human microbiome is the collection of microbial organisms that live inside of and on the surface of humans.
In this section I will review what is known ...

Archaea has been found in the gut.

Genetics.

@@

The microbiome has remarkable metabolic potential.
\cite{qin2010human}
% To understand the impact of gut microbes on human health and well-being it is crucial to assess their genetic potential. Here we describe the Illumina-based metagenomic sequencing, assembly and characterization of 3.3 million non-redundant microbial genes, derived from 576.7 gigabases of sequence, from faecal samples of 124 European individuals. The gene set, ~150 times larger than the human gene complement, contains an overwhelming majority of the prevalent (more frequent) microbial genes of the cohort and probably includes a large proportion of the prevalent human intestinal microbial genes. The genes are largely shared among individuals of the cohort. Over 99% of the genes are bacterial, indicating that the entire cohort harbours between 1,000 and 1,150 prevalent bacterial species and each individual at least 160 such species, which are also largely shared. We define and describe the minimal gut metagenome and the minimal gut bacterial genome in terms of functions present in all individuals and most bacteria, respectively.

Core microbiome \cite{turnbaugh2008core}.
Human microbiome is spatially organized, as can be seen on skin \cite{grice2009topographical}.
Bacterial community variation in human body habitats across space and time \cite{costello2009bacterial}.
Inter-individual versus intra-individual variation \cite{hmp2012structure}.

@@

Obesity

Microbial ecology: human gut microbes associated with obesity \cite{ley2006microbial}

Amish study \cite{zupancic2012analysis}
"Neither BMI nor any metabolic syndrome trait was associated with a particular gut community. Network analysis identified twenty-two bacterial species and four OTUs that were either positively or inversely correlated with metabolic syndrome traits, suggesting that certain members of the gut microbiota may play a role in these metabolic derangements."

pregancy \cite{koren2012host}

@@

Although the focus of much of the work has been on bacteria and archaea, other groups of life have attracted more attention lately, such as the fungi.
Viruses too
febrile study \cite{wylie2012sequence}

\section{Methods used to investigate the human microbiome}

Terminology: survey versus metagenome.
Definition of a single sample.
Despite their importance, I will not describe experimental techniques such as gnotobiotic mice to stay within scope.

@@

\subsection{Marker genes}
History-- bacteria?

Our knowledge of the microbial world is due to a large part derived from the work of Carl Woese and colleagues who pioneered the use of marker genes as a way to distinguish between microbial lineages.
Their work, and the scientists who followed them, used especially the 16S ribosomal gene.
This gene was chosen because...
They developed the first "universal" 16S PCR primers.
The great accomplishment of this is learning that archaea, although morphologically similar to bacteria, form their own separate lineage.

Non-culturable organisms.

While Woese and colleagues used primitive sequencing methods to gain information, modern researchers have the luxury of high throughput sequencing.
This can be

Despite the high-throughput and low cost of modern sequencing, inherent challenges remain for the use of population census by marker gene sequencing.
Most fundamentally, various microbes have different DNA extraction efficiencies, even with hard core protocols, meaning that the representation of marker gene sequences is not representative of the actual communities \cite{morganEaInVitroSimulatedMetagenome10}.
Furthermore, primer bias.
Current sequencing technology is limited to a length that is shorter than most genes, which limits the resolution of the analyses.
Different regions may be good for different groups of organisms.

16S exists in multiple copies and there can be diversity within the copies.
Kembel efforts to correct for it.

Noise and the rare biosphere?

Knowledge of the impact of microbial communities has been

\subsection{Metagenomes}

\subsection{Data generated from the various studies}
reference genomes
HMP DACC.

@@

@@

Phylogenetics is primarily used in this realm as a framework with which data can be organized.
It is common for researchers to equate 16s phylogeny with taxonomy.

@@

The reason people are attached to taxonomy is that there is a lot of information using taxonomic labels.

@@

In general, this sort of approach has led to a number projects that use phylogeny to revise taxonomy.
These attempts are less ambitious than the development of the PhyloCode, and simply work to revise the hierarchical structure of the taxonomy while (for the most part) leaving taxonomic names fixed.
Bergey's Manual of Systematic Bacteriology \cite{kreig1984bergey}.

GreenGenes.
\cite{desantis2006greengenes}
{Greengenes, a chimera-checked 16S rRNA gene database and workbench compatible with ARB}
\cite{dalevi2007automated}
Automated group assignment in large phylogenetic trees using GRUNT: GRouping, Ungrouping, Naming Tool

tax2tree.
SILVA.

@@

@@

@@

@@

\subsection{Genome-scale inquiries using phylogenetics}

GEBA \cite{wu2009phylogeny}

Sushi paper, network of gene transfer.

Reconciliation, team phyldog.

@@

NJ

@@

@@

\subsection{Software}
\cite{caporaso2010qiime}

@@

hooper2012interactions,

@@

\subsection{Notably absent}
Trait evolution methods are not common.

@@

What we can expect next.
Interactions with the immune system:
\cite{hooper2012interactions}

Clinical applications, but MALDI-TOF

Field is still not yet mature.
Need to think about study design.
